Data-Driven Customization

### 2020-08-15 paper.tex
visions

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{End-User Software Customization by Direct Manipulation of Tabular
Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{data-driven customization}, a new way
  for end users to extend software by direct manipulation without doing
  traditional programming. We augment existing user interfaces with a
  table view showing the structured data inside the application. When
  users edit the table, their changes are reflected in the original UI.
  This simple model accommodates a spreadsheet formula language and
  custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

  We illustrate the approach with Wildcard, a browser extension that
  implements data-driven customization for existing applications using
  web scraping. Through concrete examples, we show that this paradigm
  can support extensions to many real websites, ranging from sorting and
  filtering existing data to adding entire new features. We share
  reflections from our experiences using Wildcard, on both its strengths
  and limitations relative to other customization approaches. Finally,
  we explore how this paradigm might lead to new software architectures
  that encourage this form of end-user customization.
\end{abstract}

@@

data-driven customization relates to existing work in many areas. Our
goals overlap with many software customization tools, and our methods
overlap with direct manipulation interfaces for working with structured
data, including visual database query systems and spreadsheets. We
explore these connections in Section~\ref{sec:related-work}.

@@

\begin{itemize}
\tightlist
\item
  Wildcard just one instance of a broader data-driven customization

  \begin{itemize}
  \tightlist
  \item
    particular view: table
  \item
    particular context: web
  \end{itemize}
\item
  here we share some of the underlying ideas. go beyond Wildcard, room
  to grow
\end{itemize}

@@

\hypertarget{decouple-data-from-application-logic}{%
\subsection{Decouple data from application
logic}\label{decouple-data-from-application-logic}}

\begin{itemize}
\item
  thin client architecture of web has promoted a coupling of data + app

  \begin{itemize}
  \tightlist
  \item
    only way user can view the data is in the first-party tool
  \item
    APIs provide some flexibility in picking a client. but, APIs aren't
    meant for end user consumption. and, there are limits (eg, Twitter
    API restrictions)
  \item
    limited user agency over how they use their data
  \item
    compare to, say, opening a file locally
  \end{itemize}
\item
  Some people aiming towards ambitious rearchitectures in this
  space\ldots{}

  \begin{itemize}
  \tightlist
  \item
    cite local-first
  \item
    TBL Solid
  \end{itemize}
\item
  Data-driven customization: an iterative approach towards this vision;
  complementary to more radical rearchitecting.
\item
  Start with third party extraction
\item
  Integrate across apps
\item
  open the data in a different UI
\item
  Then move to first party data providing in the client (could drive
  wildcard from existing apis??)
\end{itemize}

@@

data-driven customization relates to two broad areas of related work.
Our problem statement is related to software customization tools, and
our solution approach is related to spreadsheets and other direct
manipulation interfaces.

@@

data-driven customization is most closely related to other tools that
aim to empower end users to customize software without traditional
coding.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Data-Driven Customization}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Customizing software should be as easy as using it. But most
  customization methods require a dramatic shift, from using a GUI to
  writing scripts in a programming language.

  We introduce \emph{data-driven customization}, a new way for end users
  to extend software by direct manipulation without doing traditional
  programming. We augment existing user interfaces with a table view
  showing the structured data inside the application. When users edit
  the table, their changes are reflected in the original UI. This simple
  model accommodates a spreadsheet formula language and custom data
  editing widgets, providing enough power to implement a variety of
  useful extensions.

  We illustrate the approach with Wildcard, a browser extension that
  implements data-driven customization on the web using web scraping.
  Through concrete examples, we show that this paradigm can support
  useful extensions to many real websites, and we share reflections from
  our experiences using Wildcard.

  Finally, we share our broader vision for a future where end users have
  greater access to the data in their applications, and better tools for
  taking advantage of that data in the context of everyday software
  usage.
\end{abstract}

@@

Data-driven customization relates to existing work in many areas. Our
goals overlap with many software customization tools, and our methods
overlap with direct manipulation interfaces for working with structured
data, including visual database query systems and spreadsheets. We
explore these connections in Section~\ref{sec:related-work}.

@@

We envision data-driven customization as a broad paradigm that could
extend well beyond the Wildcard proof-of-concept, and ultimately result
in new software architectures that empower end users to mold software to
their specific needs. Here we explore some of the deeper ideas
underlying our work, and future possibilities beyond the Wildcard tool.

\hypertarget{decouple-data-from-applications}{%
\subsection{Decouple data from
applications}\label{decouple-data-from-applications}}

When data is freely available outside the context of a specific
application, users have more freedom to choose a suitable application
for their needs. For example, an RSS feed can be consumed by many reader
applications---a journalist can use a power tool optimized for skimming
hundreds of news sources a day, while a casual reader can use a simple
app to keep up with a few blogs. Motivated users can even create their
own custom workflows for filtering and combining RSS feeds, either via
traditional programming or in an end-user programming environment like
\href{https://en.wikipedia.org/wiki/Yahoo!_Pipes}{Yahoo Pipes}.

However, on the Web today, data is often siloed and only accessible
through a single prescribed application. A Facebook feed can only be
viewed through the Facebook application. Podcasts, originally served
openly through RSS, are
\href{https://stratechery.com/2019/spotifys-podcast-aggregation-play/}{beginning
to become exclusive} to specific platforms like Spotify. This coupling
between data and applications leaves users at the mercy of using a
single client optimized for specific purposes (e.g., maximizing
engagement) that may not be aligned with users' individual desires.

Some services provide APIs that mitigate these siloing effects, but APIs
fail to provide a full solution to the problem. First, APIs often
provide limited access, especially when an open client ecosystem would
harm the economic incentives of a web service built on advertising in a
first-party client---in 2012, Twitter infamously
\href{https://www.theverge.com/2012/8/16/3248079/twitter-limits-app-developers-control}{imposed
restrictions} on third-party applications that mimicked the ``mainstream
Twitter consumer client experience.'' A second problem is that web APIs
have a high barrier to entry---they tend to be designed more for
programmers creating entire applications or heavy-duty automations than
for end users casually modifying their own experience.

There have been compelling suggestions for more decentralized
architectures that would give users more control of their data.
Local-first software \citep{kleppmann2019} suggests that productivity
applications should run logic and store data locally, while retaining
the benefits of realtime collaboration through peer-to-peer
synchronization. The SOLID project \citep{berners-lee2019} envisions a
decentralized future where users store data on their own servers, and
choose to grant limited access to applications. We see data-driven
customization as aligned with these decentralized visions, but
complementary to them in two ways.

\textbf{Incremental:} Rather than proposing that web services be totally
rearchitected, data-driven customization suggests a more incremental
path for adding user agency to existing software.

Data-driven customization allows for lightly augmenting a centralized
website with decentralized data storage. Section~\ref{sec:example}
demonstrated how a user's private annotations on a news site could be
stored in their browser, without needing to upload the annotation to the
website's server. A hybrid storage model is reasonable here: a
centralized storage model makes sense for most of the information on
Hacker News that is viewed by all users, but a user's private
annotations can easily just be stored in their browser. This model could
even be extended with peer-to-peer sharing---a column in a Wildcard
table could be shared directly among friends, providing shared
annotation of a common website without needing to use the website itself
as a centralized intermediary.

By using third-party data extraction, data-driven customization also
works with existing websites that do not expose structured data to the
user. Ultimately, in adversarial situations where websites are strongly
incentivized to restrict access to their data, scraping is unlikely to
be a sustainable solution. But we hypothesize that there are many more
situations where websites are \emph{neutral}: not opposed to the idea of
end user customization, but also not sufficiently motivated to create
and maintain a public API. We see these neutral situations as a context
where this kind of incremental approach could succeed. Perhaps some of
these websites might be more motivated to provide official extension
hooks if they saw the value that users were getting from unofficial
community-provided ones.

\textbf{Focus on end user customization:} Having access to the data is a
necessary but not sufficient condition for empowering end users to craft
their own software experience. Another key ingredient is providing
usable tools and interfaces for working with the data.

In data-driven customization, we focus heavily on this part of the
solution. By showing raw data in the context of a user interface and
allowing small tweaks to the original application's behavior, we provide
a smooth path for people to move from using an application to tweaking
it.

In this sense, data-driven customization is a complementary approach to
other projects that focus on getting users greater access to their data.
In a decentralized future where data is stored locally rather than in
cloud silos, interfaces like Wildcard would be one technique for
actually making use of this data in service of greater end user
flexibility.

@@

@@

Data-driven customization relates to two broad areas of related work.
Our problem statement is related to software customization tools, and
our solution approach is related to spreadsheets and other direct
manipulation interfaces.

@@

Data-driven customization is most closely related to other tools that
aim to empower end users to customize software without traditional
coding.
